---
title: "Ax example of estimation"
output: 
  html_document:
    toc: yes
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE) 
```

# Preliminary checks and extraction of the annual maxima 

We verify if the required packages have been installed.


```{r}
pkgTest <- function(x)
  {
    if (!require(x,character.only = TRUE))
    {
      install.packages(x,dep=TRUE)
        if(!require(x,character.only = TRUE)) stop("Package not found")
    }
}
pkgTest("xts")
pkgTest("lmomco")
```


```{r}
load("single-daily.Rdata")
```


Evaluation of yearly maxima

```{r}
ymax<-apply.yearly(y,FUN=max,na.rm=TRUE)
index.time <- seq(from =as.Date(index(ymax)[1]),                length.out = length(ymax),by = "year")
ymax<-ts(as.numeric(ymax),start =as.numeric(format(index(ymax),"%Y")[1]))
```

Plot of the resulting time series

```{r}
lwd <- 1.5
plot(ymax,ylab="mm",main=paste(namestation,", daily annual maxima",sep=""), col=4,lwd=lwd)
grid()
```

Missing data is not allowed in the estimation function.

```{r}
X<-as.numeric(na.omit(ymax))
X<-X[is.finite(X)]
```


# Comparison of different estimation methods

Fe fit a GEV distribution


```{r}
FF <- nonexceeds(); 
gevfitlmom<-lmr2par(X, type="gev")
gevfitmle<-mle2par(X, type="gev")
```


Plot of the intensity frequency curve


```{r}
FF <- nonexceeds(); 
gevfitmle<-mle2par(X, type="gev")
gevfitlmo<-lmr2par(X, type="gev")
plot(pp(X), sort(X), xlab="Probability",ylab="Quantile",pch="+",cex=0.6)
lines(FF, qlmomco(FF, gevfitmle), col="coral", lwd =lwd ) 
lines(FF, qlmomco(FF, gevfitlmo), col="lightblue",lwd=lwd)
legend("topleft",legend = c("MLE","L-Moments"), lwd= lwd, col=c("lightblue","coral"))
grid()
```


Plot of the intensity frequency curve

```{r}
plot(prob2T(pp(X)), sort(X), xlab="T-year return Periods",ylab="Quantile",pch="+",cex=0.6)
TT <- prob2T(FF)
lines(TT, qlmomco(FF, gevfitmle), col="lightblue", lwd =lwd)
lines(TT, qlmomco(FF, gevfitlmo), col="coral", lwd = lwd  ) 
legend("bottomright",legend = c("MLE","L-Moments"), lwd= lwd, col=c("lightblue","coral"))
grid()
```


# Parametric bootstrap 

```{r}
set.seed(18)
B <- 500
n <- length(X)
bootdata<-matrix(lmomco::rlmomco(n=B*n, gevfitmle),nrow = B)

thetahat<-function(x,est,type)
{
  est(x,type)$para 
}

quant<-function(vec,f, type) 
  {
  para<-vec2par(vec, type = type)
  as.numeric(qlmomco(f,para))
}


bootestmle<-apply(bootdata,MARGIN = 1,FUN =thetahat,est=mle2par, type="gev")

bootestlmo<-apply(bootdata,MARGIN = 1,FUN =thetahat,est=lmr2par, type="gev")
```



```{r}
alpha<-0.05
qbootmle<-apply(bootestmle,2,FUN = quant,f=FF,type ="gev")
qlimmle<-apply(qbootmle,1,FUN = quantile,prob=c(0.025,0.5,0.975))
plot(prob2T(pp(X)), sort(X), xlab="T-year return periods",ylab="Quantile",type="n",ylim=range(qlimmle))
xx <- c(TT, rev(TT))
yy <- c(qlimmle[1,],rev(qlimmle[3,]))
polygon(xx, yy,density = 30,col="lightblue",border = "transparent")
lines(TT,qlimmle[2,], col="lightblue",lwd=lwd,lty=1)
qbootlmo<-apply(bootestlmo,2,FUN = quant,f=FF,type ="gev")
qlimlmo<-apply(qbootlmo,1,FUN = quantile,prob=c(alpha/2,0.5,1-alpha/2))
yy <- c(qlimlmo[1,],rev(qlimlmo[3,]))
polygon(xx, yy,density = 30,col="coral",border = "transparent",angle=135)
lines(TT,qlimlmo[2,],col="coral",lwd=lwd,lty=2)
points(prob2T(pp(X)), sort(X),pch='+', cex=0.6)
legend("topleft", legend = c("MLE","L-Moments"), 
       fill = c( "lightblue","coral"),
       density = c(40,40), angle = c(45,135),lty=c(1,2),
       col=c( "lightblue","coral"),lwd=lwd)
title(main=paste(1-alpha,"%", "bootstrap pointwise confidence band"))
```


# Comparison of different fitted distributions

We fit the Generalized Logistic (`glo`), Generalized Normal (`gno`), and Pearson type III (`pe3`) distributions.


```{r}
glofitmle<-mle2par(X, type="glo")
gnofitmle<-mle2par(X, type="gno")
pe3fitmle<-mle2par(X, type="pe3")
```


```{r}
curve(pdfgev(x,para = gevfitmle),from = 0,to = max(X),xlab="y",ylab="f(y)",col=2,lwd=lwd,lty=1)
curve(pdfglo(x,para = glofitmle),from = 0,to = max(X),col=3,add=TRUE,lwd=lwd,lty=2)
curve(pdfgno(x,para = gnofitmle),from = 0,to = max(X),col=4,add=TRUE,lwd= lwd,lty=3)
curve(pdfpe3(x,para = pe3fitmle),from = 0,to = max(X),col=5,add=TRUE,lwd=lwd,lty=4)
grid()
legend("topright",legend = c("GEV","GLO","GNO", "PIII"), lwd= lwd, col=2:5,lty=1:4)
```


```{r}
plot(prob2T(pp(X)), sort(X), xlab="T-year return periods",ylab="Quantile",pch="+",cex=0.6)
lines(TT, qlmomco(FF, gevfitmle), col=2, lwd =lwd,lty=1)
lines(TT, qlmomco(FF, glofitmle), col=3, lwd =lwd,lty =2)
lines(TT, qlmomco(FF, gnofitmle), col=4, lwd =lwd,lty=3)
lines(TT, qlmomco(FF, pe3fitmle), col=5, lwd =lwd,lty=4)
grid()
legend("bottomright",legend = c("GEV","GLO","GNO", "PIII"), lwd= lwd, col=2:5, lty =1:4)
```

# Goodness-of-fit

Anderson–Darling tests

```{r}
set.seed(1)
gevad<-goftest::ad.test(X,null = "cdfgev",para=gevfitmle,estimated = TRUE)
gload<-goftest::ad.test(X,null = "cdfglo",para=glofitmle,estimated = TRUE)
gnoad<-goftest::ad.test(X,null = "cdfgno",para=gnofitmle,estimated = TRUE)
pe3ad<-goftest::ad.test(X,null = "cdfpe3",para=pe3fitmle,estimated = TRUE)
```


```{r}
results <- data.frame(dist =c("GEV","GLO","GNO","PE3"),
                      p.value = c(gevad$p.value,gload$p.value,gnoad$p.value,pe3ad$p.value))
```


```{r}
knitr::kable(results, digits = 3, caption = "Anderson-Darling test")
```

