---
title: "An example of IDF analysis"
output:
  html_document:
    toc: true
    df_print: paged
date: "`r Sys.Date()`"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE) 
```

# Preamble

We verify if the required packages have been installed.


```{r}
pkgTest <- function(x)
  {
    if (!require(x,character.only = TRUE))
    {
      install.packages(x,dep=TRUE)
        if(!require(x,character.only = TRUE)) stop("Package not found")
    }
}
pkgTest("IDF")
```


# Data description 

These data were used in the study "Flexible and Consistent Quantile Estimation for Intensity-Duration-Frequency Curves" (Fauer et al., 2021) See https://zenodo.org/records/5012621 for more details.

Rainfall data were collected from stations by the German Meteorological Service (DWD) and Wupperverband (corrected data). Raw time series data from the German Meteorological Service is publicly available under https://opendata.dwd.de/climate_environment/CDC/observations_germany/climate/.


# Import the data
We read the data

```{r}
yearmax<-read.csv("yearMax.csv")
metadata<-read.csv("meta.csv")
metadata<-metadata[unique(yearmax$station),]
id<-names(table(yearmax$station))[order(table(yearmax$station),decreasing = TRUE)]
id<-as.numeric(id)
sel<-id[3]
station.name<-metadata$station[metadata$num ==sel]
yearmax<-subset(yearmax,yearmax$station == sel)
```

We select the station 
`r station.name`

```{r}
d<-yearmax$ds
y<-yearmax$xdat
ymat<-matrix(yearmax$xdat,nrow = 44)
d1<-unique(d)
y1<-c(200.0, 100.0, 50.0, 20.0, 10.0, 5.0, 2.0, 1.0, 0.5)
plot(log(d),log(y),pch=20,cex=0.5, axes=FALSE,
     xlab = "Duration [h]", ylab = "Intensity [mm/h]")
box()

axis(side=1, at=log(d1),  labels=round(d1,2))
axis(side=2, at=log(y1),  labels=round(y1,1))
grid()
```


We fit model 25.10 (???) in the Chapter 25.

```{r}
library(IDF)
fit<-gev.d.fit(xdat = y,ds = d, show = FALSE)
par <- gev.d.params(fit = fit)
```


We plot the IDF curves

```{r echo=TRUE}
durations<-seq(min(d),max(d),length.out=30)
fitparams <- par
probs<-  c(0.5, 0.9, 0.99)
qs <- lapply(durations, qgev.d, p = probs, mut = fitparams[1], sigma0 = fitparams[2], xi = fitparams[3], theta = fitparams[4], eta = fitparams[5], eta2 = fitparams[6], tau = fitparams[7])
idf.array <- matrix(unlist(qs),length(probs),length(durations))
plot(log(d),log(y),pch=20,cex=0.5, axes=FALSE, 
     xlab = "Duration [h]", ylab = "Intensity [mm/h]")
box()
axis(side=1, at=log(d1),  labels=round(d1,2))
axis(side=2, at=log(y1),  labels=round(y1,1))
grid()
lwd<-1.5
lty<-1
for (i in 1:length(probs))
  lines(log(durations),log(idf.array[i,]),col=(i+1),lwd=lwd,lty=lty)
legend(x = "topright", title = "p-quantile", legend = probs,   col = (1:length(probs))+1,  lwd = lwd,lty = lty)
```

